QCL: Added value of differential correlation

### 2011-11-10 inst/doc/article.Rnw
Added the science style + updates to the paper

\usepackage{times}
\usepackage{amsmath}
\usepackage{url}
\usepackage{color}

% revise margins
\setlength{\headheight}{0.0in}
\setlength{\topmargin}{-0.25in}
\setlength{\headsep}{0.0in}
\setlength{\textheight}{9.00in}
\setlength{\footskip}{0.5in}
\setlength{\oddsidemargin}{0in}
\setlength{\evensidemargin}{0in}
\setlength{\textwidth}{6.5in}

@@

@@

@@

\begin{center}
\vspace*{1in}
\textbf{\Large QCL: Added value of differential correlation }

\bigskip \bigskip 
\bigskip \bigskip

{\large Danny Arends$^{1,2}$ and Ritsert C. Jansen$^{1}$}

\bigskip \bigskip

{\color{blue} \tt \bfseries $^{1}$ Groningen Bioinformatics Centre - University of Groningen\\
$^{2}$ Genomics Coordination Centre - University Medical Centre Groningen\\
version: \# 0.0.1\\[24pt]
received on XX XXX 201X\\
accepted on XX XXX 201X\\
}

\end{center}

\bigskip \bigskip

\textbf{\sffamily Abstract}: In this paper we present Quanitative Correlation Locus 
(QCL) mapping, a novel approach to detect genetic regulation of phenotypes in 
recombinant inbred line populations (RIL) or other experimental crosses. QCL mapping 
allows association of phenotype correlation differences to genetic markers, in almost 
the same way as QTL mapping associates genetic markers with differences in expression.
We use QCL mapping to reconstruct underlying genetic architecture, and provide insight 
into regulation of (gene) expression. QCL information in combination with classical QTL 
information leads to a more detailed insight into the genetic architecture underlying 
complex traits. 
In two example we reconstruct part of the genetic architecture in S. Cerevisiae, another 
example shows the reconstruction of the chain elongation pathway in A. Thaliana. In both 
examples we are able to map the observe differences in correlation to genetic markers, 
and show the biological relevance of QCL peeks and hotspots in regulatory network 
reconstruction.

\vfill

\noindent Corresponding Author: Danny Arends \verb|Danny.Arends@Gmail.com|
\noindent Associate Editor: XXXX XXXX
\newpage

\textbf{\sffamily Introduction}

Genetical genomics experiments have shown that gene expression variation can be mapped to 
genotypic variation (Jansen and Nap, 2001). Quantitative Trait Locus (QTL) mapping of a 
gene identifies regions in the genome for which different genotypes lead to differential 
expression level of the gene. This QTL mapping can be done on all molecular levels including, 
gene expression (eQTL), protein abundance (pQTL) and metabolites (mQTL).

Recently different methods have been proposed that allow studying, in addition to the genes 
differentially expressed between conditions, the genes that are differentially coexpressed 
between conditions (Tesson et al. 2010, Kompass et al. 2011). Such differential co-expression 
is taken as another level of information from genomic data. This level of information can for 
example be used to expose transcriptional network rewiring (Langfelder et al. 2008). Differential 
co-expression has been observed in studies of different diseases, including between healthy and 
cancer tissue studies. In parallel, correlations between phenotypes are studied for various 
reasons e.g. discovering and detecting batch effects, marker assisted breeding and causal 
inference. The importance of identifying differences in correlation has been increasingly 
recognized and this led to the development of this algorithm to perform QCL analysis in 
experimental crosses.

We proposed to map not differential co-expression but differential correlation to genetic 
variation, i.e. to identify regions in the genome for which one genotype value leads to 
correlated expression between a start gene and all other genes, exploiting the fact that any 
cis-QTL will induce variation which will in turn lead to correlation differences among 
individuals carrying the other genotype. We merge information comming from QCL and QTL to reveal 
underlying genetic architecture, which would remain hidden when using only classical QTL information.

@@

\textbf{\sffamily Methodology}
\nopagebreak

\emph{\sffamily Dataset Example 1. A. Thaliana}

Metabolite data from an A. Thaliana Landsberg erecta (Ler) and Cape Verde Islands (Cvi) 
dataset from R/qtl, 24 characterized metabolites which are the products of chain elongation. 
There are 162	recombinant inbred lines genotyped at 117 markers on 5 chromosomes. Additional 
details about the experimental settings are available in the original paper (Keurentjes et al. 2006,
Alonso-Blanco et al. 1998)

S. Cerevisiae dataset from the gene expression omnibus (GEO) (GDS1115 / GDS1116). This 
dataset contains expression profiles of parental strains and haploid progenies from a 
cross of strain BY4716 and the wild wine strain RM11-1a. Data was collected for 4482 
probes. Genotype data for the 109 progenies was available at 282 genetic markers 
obtained by contacting the authors of the original paper. Expression of probes was 
measured under growth conditions containing 1\% ethanol. Additional details about the 
experimental settings are available in the original paper (Brem et al. 2005).

@@

\textbf{\sffamily Algorithm: QCL mapping}

@@

\begin{figure}
\centering
<<singletrait, fig=TRUE, echo=FALSE, height=3.5>>=
plot(qcl_result, 2, do.legend=FALSE)
@ 
\caption{Single trait QCL profile (A. Thaliana).\label{fig:singletrait}}
\end{figure}

@@

Two things can happen: 
1) No genes show a significant change in QCL, which means we 
cannot find any interaction partners for this phenotype, or 
2) This genes has an QCL with other genes, then the algorithm adds 
these genes onto a list to scan and repeats the single phenotype 
QCL scan (step 2) for these phenotypes (not allowing for duplicate 
scans). This is repeated until no more significant QCL are found or 
a user defined search depth is reached.

@@

\bigskip
\textbf{\sffamily Results}
\nopagebreak

Global overview of QCL scan after using a full scan for QCL
<<echo=FALSE>>=
  qcl_result
@ 
Single results from a QCL mapping, shows the pattern at the highest threshold.
<<echo=FALSE>>=
  qcl_result[[2]]
@

\emph{\sffamily The Phenotypes x Marker relations}

Phenotype x Marker image plot, this is 'comparable' to a QTL scan on mutliple traits.

\begin{figure}[ht]
\centering
<<p2mmatrix, fig=TRUE, echo=FALSE, height=3.5>>=
par(mar=c(3,8,1.1,1.1))
image(qcl_result, 0.5, against="markers",do.grid=FALSE)
@ 
\caption{Phenotypes x Markers matrix (A. Thaliana).\label{fig:p2mmatrix}}
\end{figure}

\emph{\sffamily The Phenotypes x Phenotypes relations}

Phenotype x Phenotype image plot, additional information obtained by QCL mapping, 
which can be used to reconstruct the underlying genetic architecture.
\begin{figure}[ht]
\centering
<<p2pmatrix, fig=TRUE, echo=FALSE, height=3.5>>=
par(mar=c(3,8,1.1,1.1))
image(qcl_result, 0.5, against="phenotypes")
@ 
\caption{Phenotypes x Phenotypes matrix (A. Thaliana).\label{fig:p2pmatrix}}
\end{figure}

\bigskip
\textbf{\sffamily Discussion}
\nopagebreak

@@

\newpage
\textbf{\sffamily Bibliography}

Langfelder P, Horvath S (2008) WGCNA: an R package for weighted correlation network 
analysis. \emph{BMC Bioinformatics}, 9:559

Jansen R. C, and Nap J. P. (2001) Genetical Genomics: the added value from seggregation. 
\emph{Trends Genet}, 17(7), 388-391.

Tesson, B. M., Breitling R. and Jansen, R. C. (2010). DiffCoEx: a simple and sensitive method 
to find differentially coexpressed gene modules. \emph{BMC Bioinformatics}, 11,497-506.

Kompass, K. and Witte, J. (2011) - Co-regulatory expression quantitative trait loci mapping: 
method and application to endometrial cancer. \emph{BMC Medical Genomics}, 4, 6.

Broman, K.W., Wu, H., Sen, S. and Churchill, G.A. (2003) - R/qtl: QTL mapping in experimental 
crosses. \emph{Bioinformatics}, 19, 889-890.

Arends, D., Prins, P., Jansen, R.C. and Broman, K.W. (2010) - R/qtl: high-throughput multiple 
QTL mapping. \emph{Bioinformatics}, 26, 2990-2992.

R Development Core Team (2010) - R: A Language and Environment for Statistical Computing. 
http://www.R-project.org, ISBN: 3-900051-07-0

Edgar R., Domrachev M., Lash A.E. (2002) - Gene Expression Omnibus: NCBI gene expression and 
hybridization array data repository. \emph{Nucleic Acids Res.}, 30,207-210.

Keurentjes, J. J., Fu, J., de Vos, C. H., Lommen, A., Hall, R. D., Bino, R. J., van der Plas, 
L. H., Jansen, R. C., Vreugdenhil, D., Koornneef, M. (2006) - The genetics of plant metabolism. 
\emph{Nature Genetics.}, 38(7), 842-849.

Alonso-Blanco, C., Peeters, A. J., Koornneef, M., Lister, C., Dean, C., van den Bosch, N., Pot, 
J., Kuiper, M. T. (1998) - Development of an AFLP based linkage map of Ler, Col and Cvi 
Arabidopsis thaliana ecotypes and construction of a Ler/Cvi recombinant inbred line population. 
\emph{Plant J.}, 14, 259-271.

Brem RB, Storey JD, Whittle J, Kruglyak L. (2005) - Genetic interactions between polymorphisms 
that affect gene expression in yeast. \emph{Nature}, 436(7051),701-703.

\newpage
{\small
\noindent \textbf{\sffamily Package qcl function overview} \\
\noindent \begin{tabular}{ll} 
\hspace*{25mm} & \hspace*{103mm} \\ \hline
QCLscan & Genome wide scan for QCL \\
QCLscan.network & Scan for QCL using start gene(s) \\
QCLpermute & Significance thresholds for QCL by permutation \\
plot.QCL & Plot QCL profiles at increasing cut-offs \\
image.QCL & Image a multiple phenotype scan \\
\hline
\end{tabular}
}

>>>>

\usepackage{scicite}
\usepackage{times}
\usepackage{amsmath}
\usepackage{url}
\usepackage{graphicx}
\usepackage{color}

% revise margins
\setlength{\headheight}{0.0in}
\setlength{\topmargin}{0.0cm}
\setlength{\headsep}{0.0in}
\setlength{\textheight}{21.0cm}
\setlength{\footskip}{1.0cm}
\setlength{\oddsidemargin}{0in}
\setlength{\evensidemargin}{0in}
\setlength{\textwidth}{16.0cm}

@@

%The next command sets up an environment for the abstract to your paper.

\newenvironment{sciabstract}{%
\begin{quote} \bf}
{\end{quote}}

% If your reference list includes text notes as well as references,
% include the following line; otherwise, comment it out.

\renewcommand\refname{References and Notes}

% The following lines set up an environment for the last note in the
% reference list, which commonly includes acknowledgments of funding,
% help, etc.  It's intended for users of BibTeX or the {thebibliography}
% environment.  Users who are hand-coding their references at the end
% using a list environment such as {enumerate} can simply add another
% item at the end, and it will be numbered automatically.

\newcounter{lastnote}
\newenvironment{scilastnote}{%
\setcounter{lastnote}{\value{enumiv}}%
\addtocounter{lastnote}{+1}%
\begin{list}%
{\arabic{lastnote}.}
{\setlength{\leftmargin}{.22in}}
{\setlength{\labelsep}{.5em}}}
{\end{list}}

\title{QCL: Added value of differential correlation}

% Place the author information here.  Please hand-code the contact
% information and notecalls; do *not* use \footnote commands.  Let the
% author contact information appear immediately below the author names
% as shown.  We would also prefer that you don't change the type-size
% settings shown here.

\author
{Danny Arends,$^{1\ast,2}$ Ritsert C. Jansen,$^{1}$\\
\\
\normalsize{$^{1}$Groningen Bioinformatics Centre - University of Groningen}\\
\normalsize{Nijenborgh 7, 9747 AG Groningen, The Netherlands}\\
\normalsize{$^{2}$Genomics Coordination Centre - University Medical Centre Groningen}\\
\normalsize{Hanzeplein 1, 9700 RB Groningen, The Netherlands}\\
\\
\normalsize{$^\ast$To whom correspondence should be addressed; E-mail:  Danny.Arends@Gmail.com.}
}

% Include the date command, but leave its argument blank.

\date{}

%%%%%%%%%%%%%%%%% END OF PREAMBLE %%%%%%%%%%%%%%%%

@@

% Double-space the manuscript.

\baselineskip24pt

% Make the title.

\maketitle

@@

\begin{sciabstract}
In this paper we present Quanitative Correlation Locus (QCL) mapping, a novel approach 
to detect genetic regulation of phenotypes in recombinant inbred line populations (RIL) 
or other experimental crosses. QCL mapping allows association of phenotype correlation 
differences to genetic markers, in almost the same way as QTL mapping associates genetic 
markers with differences in expression.
We use QCL mapping to reconstruct underlying genetic architecture, and provide insight 
into regulation of (gene) expression. QCL information in combination with classical QTL 
information leads to a more detailed insight into the genetic architecture underlying 
complex traits. The method also provides a vital link in regulatory network reconstruction.
\end{sciabstract}

\section*{Introduction}

Genetical genomics experiments have shown that gene expression variation can be mapped to 
genotypic variation \cite{Jansen:2001}. Quantitative Trait Locus (QTL) mapping of a 
gene identifies regions in the genome for which different genotypes lead to differential 
expression level of the gene. This QTL mapping can be done on all molecular levels including, 
gene expression (eQTL), protein abundance (pQTL) and metabolites (mQTL).

Recently different methods have been proposed that allow studying, in addition to the genes 
differentially expressed between conditions, the genes that are differentially coexpressed 
between conditions \cite{Tesson:2010,Kompass:2011}. Such differential co-expression 
is taken as another level of information from genomic data. This level of information can for 
example be used to expose transcriptional network rewiring \cite{Langfelder:2008}. Differential 
co-expression has been observed in studies of different diseases, including between healthy and 
cancer tissue studies. In parallel, correlations between phenotypes are studied for various 
reasons e.g. discovering and detecting batch effects, marker assisted breeding and causal 
inference. The importance of identifying differences in correlation has been increasingly 
recognized and this led to the development of this algorithm to perform QCL analysis in 
experimental crosses.

We proposed to map not differential co-expression but differential correlation to genetic 
variation, i.e. to identify regions in the genome for which one genotype value leads to 
correlated expression between a start gene and all other genes, exploiting the fact that any 
cis-QTL will induce variation which will in turn lead to correlation differences among 
individuals carrying the other genotype. We merge information coming from QCL and QTL to reveal 
underlying genetic architecture, which would remain hidden when using only classical QTL information.

@@

\section*{Methodology}
\nopagebreak

@@

\section*{Algorithm: QCL mapping}

@@

@@

Two things can occur: 
1) No genes show a significant change in QCL, which means we 
cannot find any interaction partners for this phenotype, or 
2) This genes has an QCL with other genes, then the algorithm adds 
these genes onto a list to scan and repeats the single phenotype 
QCL scan (step 2) for these phenotypes (not allowing for duplicate 
scans). This is repeated until no more significant QCL are found or 
a user defined search depth is reached.

@@

\subsection*{Materials}
\emph{\sffamily Dataset Example 1. A. Thaliana}

Metabolite data from an A. Thaliana Landsberg erecta (Ler) and Cape Verde Islands (Cvi) 
dataset from R/qtl, 24 characterized metabolites which are the products of chain elongation. 
There are 162	recombinant inbred lines genotyped at 117 markers on 5 chromosomes. Additional 
details about the experimental settings are available in the original paper \cite{Keurentjes:2006,AlonsoBlanco:1998}

S. Cerevisiae dataset from the gene expression omnibus (GEO) \cite{Edgar:2002} (GEO 
accession GDS1115, and GDS1116, \cite{Brem:2005}. This dataset contains expression 
profiles of parental strains and haploid progenies from a cross of strain BY4716 and 
the wild wine strain RM11-1a. Data was collected for 4482 probes. Genotype data 
for the 109 progenies was available at 282 genetic markers obtained by contacting 
the authors of the original paper. 
Expression of probes was measured under growth conditions containing 1\% ethanol. 
Additional details about the experimental settings are available in the original 
paper \cite{Brem:2005}.

\bigskip
\section*{Results}
\nopagebreak

\emph{\sffamily Example 1. A. Thaliana}

QCL mapping was performed on 24 characterized metabolities measured in Arabidopsis 
Thaliana. A significance threshold for QCL mapping was determined by using 1000 
permutations, the outcomes can be seen in table XX. We opted for the stringent 
threshold at P < 0.001 to use as 'significant' QCL score for further analysis.

Overview plots of the results of the QCL mapping in Arabidopsis can be found in 
Figure XX Phenotype to Marker interaction, which is comparable to the QTL mapping 
(Fig. XX). The additional information obtained by QCL mapping is shown in a heatmap 
of the Phenotype to Phenotype interaction matrix (figure XX) and the same information 
but now represented as an interaction network is seen in figure XX
\begin{table}[ht]
\begin{tabular}{ l | c c }
P-value & QCL threshold Arabidopsis & QCL threshold Saccharomyces\\
\hline
5 \% & 0.220 & 0.435\\
1 \% & 0.294 & 0.495\\
0.1 \% & 0.375 & 0.562\\
\end{tabular}
\caption{QCL thresholds obtained after 1000 permutations}
\label{permutations}
\end{table}

\emph{\sffamily Example 1. S. Cerevisiae}

QCL mapping was performed on

\bigskip
\section*{Discussion}
\nopagebreak

@@

\section*{Acknowledgements}
This work was supported by the Centre for BioSystems Genomics (CBSG) and the 
Netherlands Consortium of Systems Biology (NCSB), both of which are part of the Netherlands 
Genomics Initiative / Netherlands Organisation for Scientific Research [to DA], and 
by the EU 7th Framework Programme under the Research Project PANACEA, [222936 to RJ].

Also we'd like to mention Joeri van der Velde and Konrad Zych for supporting me with some very 
valuable feedback on figures presented here.

We'd like to thank readers for their attention, and invite them to 
address any questions to Danny Arends, at Danny.Arends@gmail.com.

\newpage
\bibliography{scibib}
\bibliographystyle{Science}

\newpage

\section*{Figures}

\begin{figure}[ht]
  \centerline{\includegraphics{QCL.pdf}}
  \caption{QCL mapping: The underlying hypothesis of QCL mapping. Seen here is a network of 3 phenotypes, in 
  which expression of P1, P2 and P3 is measured, and we want to infer the hidden relation between P1 
  and P2. In our model a genetic pertubation is causing P1 to show a QTL at a certain marker. We expect variation 
  induced by P1 to transfer to P2, However (like in most biological systems) expression of P2 is tightly controlled, 
  and variation seen in phenotype 1 wil not induce a QTL in phenotype 2. However when we partition individuals based 
  on the genotype at the QTL marker of P1, we can make use of the correlation structure between P1 and P2. Allowing 
  us to infer the relationship between phenotype 1 and 2, when we see difference in correlation between P1A and P2A 
  and P1B and P2B. We don't observe this difference in correlation between P3 and P1, meaning there is no relation 
  between P1 and P3}
  \label{QCL_img}
\end{figure}

\begin{figure}
  \centering
<<singletrait, fig=TRUE, echo=FALSE, height=3.5>>=
plot(qcl_result, 2, do.legend=FALSE)
@ 
  \caption{Single trait QCL profile (A. Thaliana) shows the number of phenotypes that have a significant 
  QCL with the scanned phenotype at different QCL significance thresholds.\label{fig:singletrait}}
\end{figure}

\begin{figure}[ht]
  \centering
<<p2mmatrix, fig=TRUE, echo=FALSE, height=3.5>>=
par(mar=c(3,8,1.1,1.1))
image(qcl_result, 0.5, against="markers",do.grid=FALSE)
@ 
  \caption{Phenotypes x Markers matrix (A. Thaliana) this is 'comparable' to a QTL scan on multiple traits.\label{fig:p2mmatrix}}
\end{figure}

\begin{figure}[ht]
  \centering
<<p2pmatrix, fig=TRUE, echo=FALSE, height=3.5>>=
par(mar=c(3,8,1.1,1.1))
image(qcl_result, 0.5, against="phenotypes")
@ 
  \caption{Phenotypes x Phenotypes matrix (A. Thaliana) additional information obtained by QCL mapping, 
  which can be used to reconstruct the underlying genetic architecture.\label{fig:p2pmatrix}}
\end{figure}

\begin{figure}[ht]
  \centerline{\includegraphics{AraNet.pdf}}
  \caption{Reconstructed network for chain elongation in Arabidopsis Thaliana}
  \label{AraNet}
\end{figure}

\begin{figure}[ht]
  \centerline{\includegraphics{AraPerm.pdf}}
  \caption{Histogram of scores obtained durting permutation of the QCL mapping methodology}
  \label{AraPerm}
\end{figure}
